---
editor_options:
  markdown:
    wrap: 72
output:
  html_document:
    df_print: paged
---

```{r}

library(dplyr)
library(maps)
library(spdep)
library(MASS)
library(INLA)
library(inlabru)
library(sf)
library(ggplot2)
library(RColorBrewer)
library(scoringRules)


```

### Data pre-processing

```{r}
# load data

load("Agrimonia_Dataset_v_3_0_0.Rdata")

df <- AgrImOnIA_Dataset_v_3_0_0

```


```{r}
# create area_id column for spatial component (latent random effect)


# visualize the locations of our data
unique_locs <- unique(df[, c("Latitude", "Longitude")])

plot(unique_locs$Longitude, unique_locs$Latitude,
     pch = 19, col = "steelblue",
     xlab = "Longitude", ylab = "Latitude",
     main = "Monitoring Station Locations",xlim=c(5,15),ylim=c()) # the stations look pretty uniformly spread across the sampling area

map("world", region = "Italy", add = TRUE, col = "grey80")
points(unique_locs$Longitude, unique_locs$Latitude, pch = 19, col = "steelblue")

# create area_id column to use as spatial random effect
df$area_id <- as.integer(factor(paste(df$Latitude,df$Longitude,sep = "_")))


# note: animals densities are in number per square km
```


```{r}
# define a t_id temporal "season" to asses seasonal patterns (latent random effect)

df$t_id <- as.integer(factor(format(df$Time, "%Y-%m")))

# year–month temporal specification did not provide sufficient support for a structured latent temporal effect at this resolution, leading to convergence issues in INLA for the temporal component. Therefore, specify month id

df$month_id <- as.integer(format(df$Time, "%m"))


```

### Investigate target variable of interest, PM2.5 concentrations (AQ_pm25 -\> air quality pm 2.5)

```{r}

summary(df$AQ_pm25) # lots of missing PM2.5 values
sum(df$AQ_pm25 == 0, na.rm = TRUE) # a few 0 values (also imply missing as 0 concentraiton is basically impossible)

# turn 0s into NaNs
df$AQ_pm25[df$AQ_pm25 == 0] <- NaN
# remove NaNs (and 0s) to analyse distribution of PM2.5
complete_case_df <- df[!is.na(df$AQ_pm25), ]
# note: can keep NaNs in modeling bc inla still works, but discludes NaNs in likelihood calculation but still outputs posterior predictions for them which may be useful

# raw
hist(complete_case_df$AQ_pm25, freq = FALSE, main = "Raw")

# try fitting with log-normal

# fit
fit_lnorm <- fitdistr(complete_case_df$AQ_pm25, "lognormal")
meanlog   <- fit_lnorm$estimate["meanlog"]
sdlog     <- fit_lnorm$estimate["sdlog"]

# plot histogram
hist(complete_case_df$AQ_pm25, freq = FALSE,
     main = "PM2.5 with Log-Normal Overlay",
     xlab = "PM2.5 Concentration",
     col = "lightblue", border = "white",
     ylim = c(0, 0.05))

# overlay density
x_seq <- seq(0, max(complete_case_df$AQ_pm25), length.out = 500)
lines(x_seq, dlnorm(x_seq, meanlog = meanlog, sdlog = sdlog),
      col = "darkgreen", lwd = 2)
legend("topright", 
       legend = c("Observed","Log-Normal"),
       fill   = c("lightblue", NA, NA),
       lty    = c(NA, 1),
       col    = c("lightblue", "red"),
       lwd    = c(NA, 2))


# evaluate fit with AIC
AIC(fit_lnorm)


```



### Create the mesh

```{r}

statdomain <- inla.nonconvex.hull(as.matrix(unique_locs[, c("Longitude", "Latitude")]),
  convex = -0.05, concave = -0.1,
  resolution = c(100, 100))
plot(statdomain)

statmesh <- fmesher::fm_mesh_2d(boundary = statdomain,
  max.edge = c(0.2, 0.2), cutoff = 0.2)
plot(statmesh)
points(unique_locs[, c("Longitude", "Latitude")], col="red", pch=20)

```

### Fit models

```{r}

# scale covariates
df$LI_bovine_v2_sc <- as.numeric(scale(df$LI_bovine_v2))
df$LI_pigs_v2_sc <- as.numeric(scale(df$LI_pigs_v2))


# spatial points
Locations = data.frame(easting = df$Longitude, northing = df$Latitude) 
df$geometry <- sf::st_as_sf(Locations, coords = c("easting", "northing"))$geometry

# SPDE model (spde spatial effect) -> specifying the priors for spatial effect (the only-non default priors for now)
loc.spde = inla.spde2.pcmatern(
  mesh = statmesh,
  prior.range = c(1, 0.5), # this means we set P(range<1)=0.5
  prior.sigma = c(1, 0.5), # "      "   "   "  P(sigma>1)=0.5
  constr = TRUE
)

bru_opts <- list(
  # control.family = list(link = "log"), # doesnt work with lognormal
  control.compute = list(dic = TRUE, mlik = TRUE,  cpo = TRUE)
)


# M0: Fixed effects only
cmp_m0 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc

m0_bru <- bru(
  components = cmp_m0,
  family = "lognormal",
  data = df,
  options = bru_opts
)

# M1: Fixed + Temporal effect (rw1)
cmp_m1 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  f(month_id, model = "rw1", cyclic = TRUE)

m1_bru <- bru(
  components = cmp_m1,
  family = "lognormal",
  data = df,
  options = bru_opts
)

# M2: Fixed + Spatial effects (SPDE - matern)
cmp_m2 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  floc(geometry, model = loc.spde)

m2_bru <- bru(
  components = cmp_m2,
  family = "lognormal",
  data = df,
  options = bru_opts
)


# M3: Full spatio-temporal
cmp_m3 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  f(month_id, model = "rw1", cyclic = TRUE) +
  floc(geometry, model = loc.spde)

m3_bru <- bru(
  components = cmp_m3,
  family = "lognormal",
  data = df,
  options = bru_opts
)



```

### Model comparison and selection

```{r}
# model comparison and selection

# DIC
cat("M0 DIC =",  m0_bru$dic$dic, "\n")
cat("M1 DIC =",  m1_bru$dic$dic, "\n")
cat("M2 DIC =",  m2_bru$dic$dic, "\n")
cat("M3 DIC =",  m3_bru$dic$dic, "\n\n")


# NLSCPO
m0_nlscpo <- -sum(log(m0_bru$cpo$cpo), na.rm = TRUE)
m1_nlscpo <- -sum(log(m1_bru$cpo$cpo), na.rm = TRUE)
m2_nlscpo <- -sum(log(m2_bru$cpo$cpo), na.rm = TRUE)
m3_nlscpo <- -sum(log(m3_bru$cpo$cpo), na.rm = TRUE)

cat("M0 NLSCPO =", m0_nlscpo, "\n")
cat("M1 NLSCPO =", m1_nlscpo, "\n")
cat("M2 NLSCPO =", m2_nlscpo, "\n")
cat("M3 NLSCPO =", m3_nlscpo, "\n\n")


# Marginal log-likelihood
cat("M0 mlik =", m0_bru$mlik[1], "\n")
cat("M1 mlik =", m1_bru$mlik[1], "\n")
cat("M2 mlik =", m2_bru$mlik[1], "\n")
cat("M3 mlik =", m3_bru$mlik[1], "\n")


```

Although Model 3 had the best overall fit based on mlik, DIC, and NLSCP.

### Prior definitions of M3 (including "sensitivity analysis" with chosen model to decide on what priors to use)

```{r}
# Fixed effects - β (Intercept + covariates):

fit_fixed_prior <- function(prec) {
  cmp <- AQ_pm25 ~ 
    Intercept(1,
            mean.linear = 0,
            prec.linear = prec) + 
  LI_bovine_v2_sc(LI_bovine_v2_sc,
                  mean.linear = 0,
                  prec.linear = prec) +       
  LI_pigs_v2_sc(LI_pigs_v2_sc,
                mean.linear = 0,
                prec.linear = prec) +      
  f(month_id, model = "rw1", cyclic = TRUE) +
  floc(geometry, model = loc.spde)
  
  bru(
    components = cmp,
    family = "lognormal",
    data = df,
    options = bru_opts
  )
}

m3_fixed_vague  <- fit_fixed_prior(0.001) # very vague
m3_fixed_mod    <- fit_fixed_prior(1) # moderate
m3_fixed_tight  <- fit_fixed_prior(10) # tight

# Compare priors
extract_marginal <- function(model, param, label) {
  marg <- as.data.frame(model$marginals.fixed[[param]])
  marg$prior_spec <- label
  marg
}

all_params <- c("Intercept", "LI_bovine_v2_sc", "LI_pigs_v2_sc")

all_marginals <- bind_rows(
  lapply(all_params, function(p)
    extract_marginal(m3_fixed_vague, p, "Vague")),
  lapply(all_params, function(p)
    extract_marginal(m3_fixed_mod,   p, "Moderate")),
  lapply(all_params, function(p)
    extract_marginal(m3_fixed_tight, p, "Tight"))
) %>%
  mutate(param = rep(rep(all_params, each = nrow(m3_fixed_vague$marginals.fixed[[1]])), 3))

ggplot(all_marginals, aes(x = x, y = y, colour = prior_spec)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ param, scales = "free") +
  labs(title = "Prior sensitivity: fixed effects",
       x = "Coefficient value", y = "Posterior density",
       colour = "Prior spec") +
  theme_minimal()



```

The vague and weakly informative priors produced near-identical
posterior distributions, indicating that the data are sufficiently
informative to dominate prior specification across a reasonable range.
The tight prior (precision = 10, SD ≈ 0.32) produced noticeably
different posteriors, particularly for the intercept, suggesting this
specification is overly restrictive given the scale of the standardised
covariates. In the absence of strong domain knowledge regarding the
expected magnitude of livestock density effects on PM2.5, a moderate
prior (precision = 1, SD = 1) was selected for all fixed effects going
forward, providing mild regularisation while allowing the data to drive
inference.

```{r}
# Temporal effect (RW1 precision)


fit_rw1_prior <- function(u, alpha, label) {
  rw1_hyper <- list(prec = list(prior = "pc.prec", param = c(u, alpha)))
  
  cmp <- AQ_pm25 ~
    Intercept(1, mean.linear = 0, prec.linear = 1) +
    LI_bovine_v2_sc(LI_bovine_v2_sc, mean.linear = 0, prec.linear = 1) +
    LI_pigs_v2_sc(LI_pigs_v2_sc, mean.linear = 0, prec.linear = 1) +
    f(month_id, model = "rw1", cyclic = TRUE, hyper = rw1_hyper) +
    floc(geometry, model = loc.spde)
  
  bru(components = cmp, family = "lognormal",
      data = df, options = bru_opts)
}

m3_rw1_smooth <- fit_rw1_prior(u = 0.1, alpha = 0.01, label = "Smooth")    # 1% chance month-to-month jumps exceed ~10% change in PM2.5
m3_rw1_mod    <- fit_rw1_prior(u = 1,   alpha = 0.01, label = "Moderate")  # 1% chance jumps exceed ~170% change — allows real seasonal variation
m3_rw1_flex <- fit_rw1_prior(u = 10, alpha = 0.01, label = "Flexible")  # 1% chance jumps exceed 10 on log scale — no meaningful constraint on seasonal variation

# Extract posterior mean and credible interval for each month
extract_rw1 <- function(model, label) {
  rw1 <- model$summary.random$f
  rw1$prior_spec <- label
  rw1
}

rw1_sens <- bind_rows(
  extract_rw1(m3_rw1_smooth, "Smooth"),
  extract_rw1(m3_rw1_mod,    "Moderate"),
  extract_rw1(m3_rw1_flex,   "Flexible")
)

ggplot(rw1_sens, aes(x = ID, y = mean, colour = prior_spec,
                      fill = prior_spec)) +
  geom_line(linewidth = 0.9) +
  geom_ribbon(aes(ymin = `0.025quant`, ymax = `0.975quant`),
              alpha = 0.2, colour = NA) +
  labs(title = "Prior sensitivity: RW1 temporal effect",
       x = "Month", y = "Effect (log scale)",
       colour = "Prior spec", fill = "Prior spec") +
  scale_x_continuous(breaks = 1:12,
                     labels = month.abb) +
  theme_minimal()


```

All three prior specifications — smooth (u = 0.1), moderate (u = 1), and
flexible (u = 10) — produced near-identical posterior seasonal curves,
indicating that the data are sufficiently informative to determine the
shape and magnitude of the temporal effect regardless of how strongly
the prior constrains month-to-month variation. This is reassuring as it
suggests the seasonal pattern in PM2.5 is a genuine signal in the data
rather than an artefact of prior specification. Going forward, the
moderate prior (u = 1, alpha = 0.01) was selected, as it represents a
sensible default that places a weak but not negligible constraint on
temporal variation — allowing meaningful seasonal fluctuation while
providing mild regularisation — without requiring strong domain
knowledge about the exact smoothness of the seasonal cycle.

```{r}
# Spatial effect (SPDE: ρ, σ)

fit_spde_prior <- function(range0, Prange, sigma0, Psigma, label) {
  spde <- inla.spde2.pcmatern(
    mesh        = statmesh,
    prior.range = c(range0, Prange),
    prior.sigma = c(sigma0, Psigma),
    constr=TRUE
  )
  cmp <- AQ_pm25 ~
    Intercept(1, mean.linear = 0, prec.linear = 1) +
    LI_bovine_v2_sc(LI_bovine_v2_sc, mean.linear = 0, prec.linear = 1) +
    LI_pigs_v2_sc(LI_pigs_v2_sc, mean.linear = 0, prec.linear = 1) +
    f(month_id, model = "rw1", cyclic = TRUE,
      hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01)))) +
    floc(geometry, model = spde)
  
  bru(components = cmp, family = "lognormal",
      data = df, options = bru_opts)
}

m3_spde_vague  <- fit_spde_prior(0.1, 0.5, 5, 0.5, "Vague") # range median ~10km, large spatial variance allowed variation
m3_spde_mod    <- fit_spde_prior(1,   0.5, 1,   0.5, "Moderate") # range median ~100km, roughly a third of domain width
m3_spde_tight  <- fit_spde_prior(3,   0.9, 0.5, 0.5, "Tight") # 90% chance range < 300km, modest spatial variance
# Psigma = 0.5 ->  treat sigma0 as the prior median for sigma in each scenario


extract_spde_marginal <- function(model, param_name, label) {
  marg <- as.data.frame(
    inla.smarginal(model$marginals.hyperpar[[param_name]])
  )
  marg$prior_spec <- label
  marg
}

range_sens <- bind_rows(
  extract_spde_marginal(m3_spde_vague, "Range for floc", "Vague"),
  extract_spde_marginal(m3_spde_mod,   "Range for floc", "Moderate"),
  extract_spde_marginal(m3_spde_tight, "Range for floc", "Tight")
)

ggplot(range_sens, aes(x = x, y = y, colour = prior_spec)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Sensitivity: SPDE range",
       x = "Range (degrees)", y = "Posterior density",
       colour = "Prior spec") +
  theme_minimal()

sigma_sens <- bind_rows(
  extract_spde_marginal(m3_spde_vague, "Stdev for floc", "Vague"),
  extract_spde_marginal(m3_spde_mod,   "Stdev for floc", "Moderate"),
  extract_spde_marginal(m3_spde_tight, "Stdev for floc", "Tight")
)

ggplot(sigma_sens, aes(x = x, y = y, colour = prior_spec)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Sensitivity: SPDE sigma",
       x = "Sigma", y = "Posterior density",
       colour = "Prior spec") +
  theme_minimal()




```

There is greater prior sensitivity than observed for the fixed effects
or temporal component, particularly for the range parameter. The vague
prior allowed the range to collapse toward very short distances
inconsistent with the spatial scale of the domain, while the moderate
and tight priors produced more stable and interpretable posteriors. For
sigma, all three specifications concentrated around similar values
(\~0.8–0.9), however the moderate prior produced a notably tighter
posterior, suggesting the moderate sigma prior (median = 1) is
well-aligned with the signal in the data, while the vague and tight
specifications introduced unnecessary additional uncertainty. Given the
absence of strong domain knowledge, the moderate prior (range median
\~100km, sigma median = 1) was selected as it produced well-identified
posteriors for both parameters without unduly constraining the spatial
field, and reflects a reasonable belief that spatial correlation in
PM2.5 operates at a regional rather than hyper-local scale across the
study domain

```{r}
# Log-normal likelihood precision

fit_likelihood_prior <- function(u, alpha, label) {
  spde <- inla.spde2.pcmatern(
    mesh        = statmesh,
    prior.range = c(1, 0.5),
    prior.sigma = c(1, 0.5),
    constr      = TRUE
  )
  cmp <- AQ_pm25 ~
    Intercept(1, mean.linear = 0, prec.linear = 1) +
    LI_bovine_v2_sc(LI_bovine_v2_sc, mean.linear = 0, prec.linear = 1) +
    LI_pigs_v2_sc(LI_pigs_v2_sc, mean.linear = 0, prec.linear = 1) +
    f(month_id, model = "rw1", cyclic = TRUE,
      hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01)))) +
    floc(geometry, model = spde)
  bru(
    components = cmp,
    family     = "lognormal",
    data       = df,
    options    = list(
      control.family  = list(
        hyper = list(
          prec = list(prior = "pc.prec", param = c(u, alpha))
        )
      ),
      control.compute = list(waic = TRUE, cpo = FALSE, config = TRUE)
    )
  )
}


# posterior precision ~2.67 → residual SD ≈ 1/sqrt(2.67) ≈ 0.61
m3_bru$summary.hyperpar["Precision for the lognormal observations", ]

m3_lik_vague  <- fit_likelihood_prior(u = 10,  alpha = 0.01, label = "Vague")    # essentially flat — well above posterior SD of 0.61
m3_lik_mod    <- fit_likelihood_prior(u = 1,   alpha = 0.01, label = "Moderate") # mild constraint, brackets posterior from above
m3_lik_tight  <- fit_likelihood_prior(u = 0.3, alpha = 0.01, label = "Tight")    # deliberately below posterior SD — tests if data dominate


# extract marginals
extract_lik_marginal <- function(model, label) {
  marg <- as.data.frame(
    inla.smarginal(
      model$marginals.hyperpar[["Precision for the lognormal observations"]]
    )
  )
  marg$prior_spec <- label
  marg
}

lik_sens <- bind_rows(
  extract_lik_marginal(m3_lik_vague,  "Vague"),
  extract_lik_marginal(m3_lik_mod,    "Moderate"),
  extract_lik_marginal(m3_lik_tight,  "Tight")
)

# plot
ggplot(lik_sens, aes(x = x, y = y, colour = prior_spec)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Sensitivity: lognormal likelihood precision",
       x = "Precision", y = "Posterior density",
       colour = "Prior spec") +
  theme_minimal()

```


All three prior specifications — vague (u = 10), moderate (u = 1), and
tight (u = 0.3) — produced near-identical posterior distributions,
indicating that the data are sufficiently informative about the residual
variance to dominate prior specification across the range tested. Given
this robustness, the moderate prior (u = 1) was selected going forward
as it provides mild regularisation without being overly restrictive, and
reflects a reasonable belief that residual variation in log PM2.5 is
unlikely to exceed 1 on the log scale while still allowing the data to
drive the posterior entirely.


All three prior specifications — vague (u = 10), moderate (u = 1), and tight (u = 0.3) — produced near-identical posterior distributions, indicating that the data are sufficiently informative about the residual variance to dominate prior specification across the range tested. Given this robustness, the moderate prior (u = 1) was selected going forward as it provides mild regularisation without being overly restrictive, and reflects a reasonable belief that residual variation in log PM2.5 is unlikely to exceed 1 on the log scale while still allowing the data to drive the posterior entirely.



```{r}
# Model with select priors (mean.linear and prec.linear are priors, all the others are hyperpriors)

# final SPDE with chosen priors
loc.spde <- inla.spde2.pcmatern(
  mesh        = statmesh,
  prior.range = c(1, 0.5),
  prior.sigma = c(1, 0.5),
  constr      = TRUE
)

cmp_m3 <- AQ_pm25 ~
  Intercept(1,
            mean.linear = 0, prec.linear = 1) +
  LI_bovine_v2_sc(LI_bovine_v2_sc,
                  mean.linear = 0, prec.linear = 1) +
  LI_pigs_v2_sc(LI_pigs_v2_sc,
                mean.linear = 0, prec.linear = 1) +
  f(month_id, model = "rw1", cyclic = TRUE,
    hyper = list(prec = list(prior = "pc.prec",
                             param = c(1, 0.01)))) +
  floc(geometry, model = loc.spde)

m3_revised <- bru(
  components = cmp_m3,
  family     = "lognormal",
  data       = df,
  options    = list(
    control.family  = list(
      hyper = list(
        prec = list(prior = "pc.prec", param = c(1, 0.01))
      )
    ),
    control.compute = list(cpo = TRUE, dic=TRUE, config = TRUE, mlik=TRUE)
  )
)

```

```{r}
# compare M3 against M3 Revised (for sensitivity analysis purposes)

# DIC
cat("M3 DIC =",  m3_bru$dic$dic, "\n")
cat("M3 Revised DIC =",  m3_revised$dic$dic, "\n\n")

# NLSCPO
m3_nlscpo <- -sum(log(m3_bru$cpo$cpo), na.rm = TRUE)
m3_nlscpo_revised <- -sum(log(m3_bru$cpo$cpo), na.rm = TRUE)

cat("M3 NLSCPO =", m3_nlscpo, "\n")
cat("M3 NLSCPO =", m3_nlscpo_revised, "\n\n")

# Marginal log-likelihood
cat("M3 mlik =", m3_bru$mlik[1], "\n")
cat("M3 Revised mlik =", m3_revised$mlik[1], "\n")



```


Remarks: -All priors were chosen empirically through sensitivity
analysis rather than arbitrarily -The moderate specification was
selected consistently across all components, reflecting a principled
philosophy of mild regularisation without domain knowledge -Where
posteriors were robust across specs (fixed effects, RW1, likelihood
precision), the moderate prior is defensible simply as a sensible
default -Where posteriors were sensitive (SPDE range and sigma), the
moderate prior was selected for producing the most stable and
interpretable posteriors



Remarks:
-Priors were selected through empirical sensitivity analysis rather than arbitrarily, with a consistent preference for moderate regularisation across all components.
-Fixed effects, the RW1 term, and likelihood precision were robust to prior specification, while the SPDE parameters showed expected sensitivity; the moderate prior was retained for stability and interpretability.
-Model diagnostics indicate negligible impact of prior revision on performance: DIC is unchanged (736102.1 vs 736101.9), NLSCPO is identical (365493.7), and marginal likelihood shows a slight improvement (-368296.5 vs -368277.8).
-Overall, the model is robust to reasonable prior perturbations, supporting the chosen specification as stable for inference and prediction.



### Global diagnostics of chosen model

```{r}
# sample from posterior

df_obs <- df[!is.na(df$AQ_pm25), ] # disclude rows with missing response (want to compare observed against predicted)

df_obs |>
    group_by(IDStations) |>
    summarise(na_bovine = sum(is.na(LI_bovine_v2_sc))) |>
    filter(na_bovine > 0) # there is just one station with no livestock data, so remove that station (there are no rows in df_obs with missing location or month id)
df_obs <- df_obs |> filter(IDStations != "STA-CH0033A") # remove station bc all its livestock data is missing

n_sims <- 1000
y_rep <- matrix(NA, nrow = nrow(df_obs), ncol = n_sims)
post_samples <- inla.posterior.sample(n_sims, m3_revised)

for(i in 1:n_sims){
  s <- post_samples[[i]]
  eta <- s$latent[1:nrow(df_obs)]        # linear predictor for each row
  sigma <- 1 / sqrt(s$hyperpar[[1]])     # lognormal sd from precision hyperpar
  y_rep[, i] <- rlnorm(nrow(df_obs), meanlog = eta, sdlog = sigma)
}

dim(y_rep)
y_rep_all <- as.vector(y_rep)

```


```{r}
# bayesian studentised residuals

# compute residuals
n <- nrow(df_obs)
y <- log(df_obs$AQ_pm25)

X <- cbind(1, df_obs$LI_bovine_v2_sc, df_obs$LI_pigs_v2_sc)
XtX_inv <- solve(t(X) %*% X)
h <- rowSums((X %*% XtX_inv) * X)

studentisedred <- matrix(NA, nrow = n, ncol = n_sims)
for(l in 1:n_sims){
  s <- post_samples[[l]]
  eta <- s$latent[1:n]                   
  sigma <- 1 / sqrt(s$hyperpar[[1]])
  studentisedred[,l] <- (y - eta) / sqrt(sigma^2 * (1 - h))
}
studentisedredm <- apply(studentisedred, 1, mean)


# qq plot - checks log normality assumption
qqnorm(studentisedredm,
       main="",
       xlab = "Theoretical Quantiles",
       ylab = "Sample Quantiles",
       pch = 16, cex = 0.4, col = "black",
       xlim = c(-3,3), ylim = c(-3,3),
       cex.lab = 1.25)
qqline(studentisedredm, col = "red", lwd = 1.5)


# residuals against index plot - checks independence of error terms
# plot(seq_along(studentisedredm), studentisedredm, xlab="Index", ylab="Bayesian studentised residual (posterior mean)")

# look at the extreme observation
df_obs[which(studentisedredm < -10), c("IDStations", "Time", "AQ_pm25", "LI_bovine_v2_sc", "LI_pigs_v2_sc")] # dont plot extreme outlier (this is the value with the smallest pm25 observation, way smaller than any others, basically 0)

plot(seq_along(studentisedredm), studentisedredm, 
     xlab = "Index", 
     ylab = "Bayesian studentised residual (posterior mean)", 
     ylim = c(-8, 8),
     pch = 16,      
     cex = 0.2)

# residuals against posterior mean of fitted values - linearity and constant variance of residuals
fittedvalues <- matrix(0, nrow = n, ncol = n_sims)
for(i in 1:n_sims){
  s <- post_samples[[i]]
  eta <- s$latent[1:n]
  fittedvalues[, i] <- exp(eta)
}

fittedvaluesm<-numeric(n)
for(i in 1:n){fittedvaluesm[i]=mean(fittedvalues[i,])}
plot(fittedvaluesm,studentisedredm,
     xlab="Fitted value (posterior mean)",
     ylab="Bayesian Studentised residual (posterior mean)",
     ylim=c(-8, 8), # get rid of that same outlier here too
     pch = 16,
     cex=0.2)


```



```{r}
# histogram of sampled stats and observed ones

df_obs <- df_obs[which(df_obs$AQ_pm25 > 3.699e-20), ] # get rid of that extreme outlier

yrepmin <- apply(y_rep, 2, min)
yrepmax <- apply(y_rep, 2, max)
yrepmedian <- apply(y_rep, 2, median)
require(fBasics)
yrepskewness <- apply(y_rep, 2, skewness)

obs_min <- min(df_obs$AQ_pm25)
obs_max <- max(df_obs$AQ_pm25)
obs_median <- median(df_obs$AQ_pm25)
obs_skewness <- skewness(df_obs$AQ_pm25)

# plot
par(mfrow = c(2, 2))

hist(yrepmin, main = "Min", xlab = "Min PM2.5", col = "lightgrey", border = "white")
abline(v = obs_min, col = "red", lwd = 2)

hist(yrepmax, main = "Max", xlab = "Max PM2.5", col = "lightgrey", border = "white")
abline(v = obs_max, col = "red", lwd = 2)

hist(yrepmedian, main = "Median", xlab = "Median PM2.5", col = "lightgrey", border = "white")
abline(v = obs_median, col = "red", lwd = 2)

hist(yrepskewness, main = "Skewness", xlab = "Skewness PM2.5", col = "lightgrey", border = "white",  xlim = c(min(obs_skewness), 16))
abline(v = obs_skewness, col = "red", lwd = 2)

par(mfrow = c(1, 1))




```


```{r}
# plot of observed vs predicted values

pred_mean <- rowMeans(y_rep)

# plot
hist(df_obs$AQ_pm25, 
     freq     = FALSE,
     main     = "",
     xlab     = "PM2.5",
     col      = rgb(0, 0, 1, 0.3),
     border   = "white",
     breaks   = 50,
     ylim = c(0, 0.07))

lines(density(pred_mean), col = "red", lwd = 2)

legend("topright",
       legend = c("Observed", "Posterior predictive mean"),
       fill   = c(rgb(0, 0, 1, 0.3), NA),
       border = c("white", NA),
       lty    = c(NA, 1),
       col    = c(NA, "red"),
       lwd    = c(NA, 2))






```




###  Convergence Diagnostics

```{r}

# ??!!




```



### Visualise the random effects

```{r}

# Zac's
 meshlocs <- as.data.frame(statmesh$loc); names(meshlocs) <- c("lon","lat")
# 
# # (M1) Temporal random effect
# plot(m1_bru$summary.random$f$mean, type="l")
# 
# # (M3) Temporal random effect
# plot(m3_bru$summary.random$ftime$mean, type="l")
# 
# # (M2) Spatial random effect
# ggplot(meshlocs, aes(x = lon, y = lat, color = m2_bru$summary.random$floc$mean)) +
#   geom_point(size = 2) +
#   scale_color_viridis_c() +
#   coord_equal() +
#   theme_minimal()
# 
# # (M3) Spatial random effect
# ggplot(meshlocs, aes(x = lon, y = lat, color = m3_bru$summary.random$floc$mean)) +
#   geom_point(size = 2) +
#   scale_color_viridis_c() +
#   coord_equal() +
#   theme_minimal()


# Jackson's

# temporal (M3)
plot(m3_revised$summary.random$f$mean, type = "l",
     main = "M3: Temporal Random Effect",
     xlab = "Time index",
     ylab = "Effect (mean)")

# spatial (M3)
ggplot(meshlocs, aes(x = lon, y = lat, color = m3_revised$summary.random$floc$mean)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  coord_equal() +
  theme_minimal() +
  labs(title = "M3: Spatial Random Effect",
       color = "Effect")


# Zac's updated (spatial only)

pix <- fm_pixels(statmesh, dims = c(200, 200),format = "sf")
pred <- predict(m3_revised, pix, formula=~floc)
ggplot() +
  gg(statmesh) +
  gg(pred, aes(color = mean))+
  scale_color_gradientn(colours=brewer.pal(11, "RdYlBu"),
                        limits=range(pred$mean))+
  geom_point(
  data = df,
  aes(x = Longitude, y = Latitude,color = area_id == 108))
  

# how to generate things from existing datapoints using predict
pred2 <- predict(m3_revised,df[,c("geometry","LI_bovine_v2_sc","LI_pigs_v2_sc","month_id")], 
                 formula=AQ_pm25~Intercept+LI_bovine_v2_sc+LI_pigs_v2_sc+f+floc)

```

**Fixed effect results**

```{r}

post_bovine <- m3_revised$marginals.fixed$LI_bovine_v2_sc
plot(post_bovine, type="l", main="Posterior marginal for bovine FE",
     ylab="Posterior density", xlab="FE value")

post_pig <- m3_revised$marginals.fixed$LI_pigs_v2_sc
plot(post_pig, type="l", main="Posterior marginal for pig FE",
     ylab="Posterior density", xlab="FE value")

intercept <- m3_revised$marginals.fixed$Intercept
plot(intercept, type="l", main="Posterior marginal for intercept",
     ylab="Posterior density", xlab="FE value")


```

The posterior distribution for the intercept is centred around 2.65 and
contains almost all probability between 2.2 and 3.0. The fixed effect
coefficients for the affect of the density of cows and pigs both have
all their mass below zero.

Due to the log link function, for fixed effects exp(coef) act as
multipliers. That is, if the density of bovines increases by one unit,
the PM2.5 reading increases by a factor of exp(coef_bovine).

```{r}

multipliers <- exp(m3_bru$summary.fixed[,c(3,4,5)])
rownames(multipliers) <- c("Intercept","Bovine","Pigs")
multipliers



```

The table displays the median effect, plus quantiles giving a 95%
credible interval for the fixed effect parameter. The multiplier for
both animals is below 1, suggesting that the presence of more animals at
a given location is associated with a decrease in the pollution at that
location.

