---
editor_options:
  markdown:
    wrap: 72
output:
  html_document:
    df_print: paged
---

```{r}

library(dplyr)
library(maps)
library(spdep)
library(MASS)
library(INLA)
library(inlabru)
library(sf)
library(ggplot2)

```


### Data pre-processing

```{r}
# load data

load("Agrimonia_Dataset_v_3_0_0.Rdata")

df <- AgrImOnIA_Dataset_v_3_0_0

```



```{r}
# create area_id column for spatial component (latent random effect)


# visualize the locations of our data
unique_locs <- unique(df[, c("Latitude", "Longitude")])

plot(unique_locs$Longitude, unique_locs$Latitude,
     pch = 19, col = "steelblue",
     xlab = "Longitude", ylab = "Latitude",
     main = "Monitoring Station Locations") # the stations look pretty uniformly spread across the sampling area

map("world", region = "Italy", add = TRUE, col = "grey80")
points(unique_locs$Longitude, unique_locs$Latitude, pch = 19, col = "steelblue")

# create area_id column to use as spatial random effect
df$area_id <- as.integer(factor(paste(df$Latitude,df$Longitude,sep = "_")))


```




```{r}
# define a t_id temporal "season" to asses seasonal patterns (latent random effect)

df$t_id <- as.integer(factor(format(df$Time, "%Y-%m")))

# year–month temporal specification did not provide sufficient support for a structured latent temporal effect at this resolution, leading to convergence issues in INLA for the temporal component. Therefore, specify month id

df$month_id <- as.integer(format(df$Time, "%m"))


```



### Investigate target variable of interest, PM2.5 concentrations (AQ_pm25 -> air quality pm 2.5)

```{r}

summary(df$AQ_pm25) # lots of missing PM2.5 values
sum(df$AQ_pm25 == 0, na.rm = TRUE) # a few 0 values (also imply missing as 0 concentraiton is basically impossible)

# turn 0s into NaNs
df$AQ_pm25[df$AQ_pm25 == 0] <- NaN
# remove NaNs (and 0s) to analyse distribution of PM2.5
complete_case_df <- df[!is.na(df$AQ_pm25), ]
# note: can keep NaNs in modeling bc inla still works, but discludes NaNs in likelihood calculation but still outputs posterior predictions for them which may be useful

# raw
hist(complete_case_df$AQ_pm25, freq = FALSE, main = "Raw")

# try fitting with log-normal

# fit
fit_lnorm <- fitdistr(complete_case_df$AQ_pm25, "lognormal")
meanlog   <- fit_lnorm$estimate["meanlog"]
sdlog     <- fit_lnorm$estimate["sdlog"]

# plot histogram
hist(complete_case_df$AQ_pm25, freq = FALSE,
     main = "PM2.5 with Log-Normal Overlay",
     xlab = "PM2.5 Concentration",
     col = "lightblue", border = "white",
     ylim = c(0, 0.05))

# overlay density
x_seq <- seq(0, max(complete_case_df$AQ_pm25), length.out = 500)
lines(x_seq, dlnorm(x_seq, meanlog = meanlog, sdlog = sdlog),
      col = "darkgreen", lwd = 2)
legend("topright", 
       legend = c("Observed","Log-Normal"),
       fill   = c("lightblue", NA, NA),
       lty    = c(NA, 1),
       col    = c("lightblue", "red"),
       lwd    = c(NA, 2))


# evaluate fit with AIC
AIC(fit_lnorm)


```

### Create the mesh


```{r}

statdomain <- inla.nonconvex.hull(as.matrix(unique_locs[, c("Longitude", "Latitude")]),
  convex = -0.05, concave = -0.1,
  resolution = c(100, 100))
plot(statdomain)

statmesh <- fmesher::fm_mesh_2d(boundary = statdomain,
  max.edge = c(0.2, 0.2), cutoff = 0.2)
plot(statmesh)
points(unique_locs[, c("Longitude", "Latitude")], col="red", pch=20)

```
### Fit models

```{r}

# scale covariates
df$LI_bovine_v2_sc <- as.numeric(scale(df$LI_bovine_v2))
df$LI_pigs_v2_sc <- as.numeric(scale(df$LI_pigs_v2))


# spatial points
Locations = data.frame(easting = df$Longitude, northing = df$Latitude) 
df$geometry <- sf::st_as_sf(Locations, coords = c("easting", "northing"))$geometry

# SPDE model (spde spatial effect) -> specifying the priors for spatial effect (the only-non default priors for now)
loc.spde = inla.spde2.pcmatern(
  mesh = statmesh,
  prior.range = c(1, 0.5),
  prior.sigma = c(1, 0.5),
  constr = TRUE
)

bru_opts <- list(
  # control.family = list(link = "log"), # doesnt work with lognormal
  control.compute = list(waic = TRUE, cpo = FALSE, config = TRUE)
)


# M0: Fixed effects only
cmp_m0 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc

m0_bru <- bru(
  components = cmp_m0,
  family = "lognormal",
  data = df,
  options = bru_opts
)

# M1: Fixed + Temporal effect (rw1)
cmp_m1 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  f(month_id, model = "rw1", cyclic = TRUE)

m1_bru <- bru(
  components = cmp_m1,
  family = "lognormal",
  data = df,
  options = bru_opts
)

# M2: Fixed + Spatial effects (SPDE - matern)
cmp_m2 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  floc(geometry, model = loc.spde)

m2_bru <- bru(
  components = cmp_m2,
  family = "lognormal",
  data = df,
  options = bru_opts
)


# M3: Full spatio-temporal
cmp_m3 <- AQ_pm25 ~ 
  Intercept(1) + 
  LI_bovine_v2_sc +
  LI_pigs_v2_sc + 
  f(month_id, model = "rw1", cyclic = TRUE) +
  floc(geometry, model = loc.spde)

m3_bru <- bru(
  components = cmp_m3,
  family = "lognormal",
  data = df,
  options = bru_opts
)



```


### Model comparison and selection


```{r}
# model comparison and selection

# WAIC -> balanced model fit and complexity
cat("M0 WAIC =",  m0_bru$waic$waic, "\n")
cat("M1 WAIC =",  m1_bru$waic$waic, "\n")
cat("M2 WAIC =",  m2_bru$waic$waic, "\n")
cat("M3 WAIC =",  m3_bru$waic$waic)

# -> select full spatio-temporal model
```

### Prior definitions of M3 (including "sensitivity analysis" with chosen model to decide on what priors to use)

```{r}
# β (Intercept + covariates):
# beta_priors <- list(
#   A = list(mean = 0, prec = 1e-6),
#   B = list(mean = 0, prec = 1e-3),
#   C = list(mean = 0, prec = 1e-2))
# 
# models <- list()
# for (p in names(beta_priors)) {
#   beta_prior <- beta_priors[[p]]
#   cmp <- AQ_pm25 ~
#     Intercept(1, prior = beta_prior) +
#     LI_bovine_v2_sc(prior = beta_prior) +
#     LI_pigs_v2_sc(prior = beta_prior) +
#     f(month_id, model = "rw1", cyclic=TRUE) +
#     floc(geometry, model = loc.spde)
#   models[[p]] <- bru(
#     components = cmp,
#     family = "gamma",
#     data = df
#   )
# }


```


```{r}
# Temporal effect (RW1 precision)






```



```{r}
# Spatial effect (SPDE: ρ, σ)




```



```{r}
# Log-normal likelihood precision



```



```{r}
# Joint sensitivity


```


# Visualise the random effects

```{r}

# Zac's
# meshlocs <- as.data.frame(statmesh$loc); names(meshlocs) <- c("lon","lat")
# 
# # (M1) Temporal random effect
# plot(m1_bru$summary.random$f$mean, type="l")
# 
# # (M3) Temporal random effect
# plot(m3_bru$summary.random$ftime$mean, type="l")
# 
# # (M2) Spatial random effect
# ggplot(meshlocs, aes(x = lon, y = lat, color = m2_bru$summary.random$floc$mean)) +
#   geom_point(size = 2) +
#   scale_color_viridis_c() +
#   coord_equal() +
#   theme_minimal()
# 
# # (M3) Spatial random effect
# ggplot(meshlocs, aes(x = lon, y = lat, color = m3_bru$summary.random$floc$mean)) +
#   geom_point(size = 2) +
#   scale_color_viridis_c() +
#   coord_equal() +
#   theme_minimal()


# Jackson's

# temporal (M3)
plot(m3_bru$summary.random$f$mean, type = "l",
     main = "M3: Temporal Random Effect",
     xlab = "Time index",
     ylab = "Effect (mean)")

# spatial (M3)
ggplot(meshlocs, aes(x = lon, y = lat, color = m3_bru$summary.random$floc$mean)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  coord_equal() +
  theme_minimal() +
  labs(title = "M3: Spatial Random Effect",
       color = "Effect")

```



### Convergence diagnostics

```{r}







```


### Global diagnostics of chosen model (CRPS, PPP values)

```{r}





```






